---
title: "Practice Materials for Exam 1"
subtitle: ""
author: "Ranadeep Daw"
date: "2024-09-07"
output: 
  revealjs::revealjs_presentation:
    theme: solar
    toc: true
    toc_depth: 2
    center: true
    reveal_options:
      slideNumber: 'c/t'
      previewLinks: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = T, eval = T)
```



## Paired or Independent?


* If data for one subject is present in both the samples, it is **paired test**.

* If one subject belongs to exactly one of the two group, it is an **independent test**.  


## Question 1

* **Problem:**
A group of 10 people are given a new diet plan. Their weight is measured before and after the diet. Test the hypothesis that the new diet plan will lead to a significant weight loss.



* **Paired or independent**?




## Question 1

* **Problem:**
A group of 10 people are given a new diet plan. Their weight is measured before and after the diet. Test the hypothesis that the new diet plan will lead to a significant weight loss.



* **Paired or independent**?

* <span style = "color:red"> Paired </span>
* Each subject or person is measured both before and after the diet.






## Question 2

* **Problem:**
A group of students takes a standardized test both before and after the implementation of a new teaching method. Test the hypothesis that the new method improves their scores.


* **Paired or independent**?


## Question 2

* **Problem:**
A group of students takes a standardized test both before and after the implementation of a new teaching method. Test the hypothesis that the new method improves their scores.


* **Paired or independent**?

* <span style = "color:red"> Paired </span>
* Each subject or person is taking exam both before and after the implementation of the new teaching method.




## Question 3


* **Problem:**
There are two groups of patients. Group 1 are given different a generic drug; group 2 receives the branded one. Test the hypothesis that the both the drugs have similar effect on the patients.


* **Paired or independent**?


## Question 3


* **Problem:**
There are two groups of patients. Group 1 are given different a generic drug; group 2 receives the branded one. Test the hypothesis that the both the drugs have similar effect on the patients.



* **Paired or independent**?

* <span style = "color:red"> Independent </span>
* Each subject or person receives only one drug. 





## Question 4

* **Problem:** Two groups of plants are given different fertilizers. The height of the plants is measured after a month.

* **Hypothesis:** Fertilizer A will lead to taller plants than Fertilizer B.



* **Paired or independent**?



## Question 4

* **Problem:** Two groups of plants are given different fertilizers. The height of the plants is measured after a month.

* **Hypothesis:** Fertilizer A will lead to taller plants than Fertilizer B.



* **Paired or independent**?

* <span style = "color:red"> Independent </span>
* Each subject or plant gets one fertilizer only.



## Question 5

* **Problem:** Employees at UWF are surveyed about their job satisfaction. Their responses are compared. Test that employees at Biology departments will report higher job satisfaction than employees at Math department.



* **Paired or independent**?


## Question 5

* **Problem:** Employees at UWF are surveyed about their job satisfaction. Their responses are compared. Test that employees at Biology departments will report higher job satisfaction than employees at Math department.



* **Paired or independent**?

* <span style = "color:red"> Independent </span>
* Each subject or employee is either in the Biology dept, or in the Math dept.






## Question 6

* **Problem:** Employees at UWF are surveyed each year about their job satisfaction. Their responses are compared. Test that employees reported a higher job satisfaction in 2023 compared to the one in 2022.



* **Paired or independent**?


## Question 5

* **Problem:** Employees at UWF are surveyed each year about their job satisfaction. Their responses are compared. Test that employees reported a higher job satisfaction in 2023 compared to the ones in 2022.



* **Paired or independent**?

* <span style = "color:red"> Paired </span>
* Each subject or employee is recording the data twice; once in 2022 and once in 2023.



## Possible hypothesis

* Here $\mu_1$, $\mu_2$ are population means for the two samples. $M_1$, $M_2$ are population medians for the two samples.
<br>


* Case 1: $H_0: \mu_1 - \mu_2 = 0$; $H_A: \mu_1 - \mu_2 \neq 0$.
* Case 2: $H_0: \mu_1 - \mu_2 \leq 0$; $H_A: \mu_1 - \mu_2 > 0$.
* Case 3: $H_0: \mu_1 - \mu_2 \geq 0$; $H_A: \mu_1 - \mu_2 < 0$.






## Parametric vs Nonparametric


* So far we have done only the following:

  * **Parametric tests**: t tests. One sample and two samples.
    - Example of parametric case: Normality assumption.
  
  * **Nonparametric**: sign test, Wilcoxon rank sum, Wilcoxon signed rank.
  
  
## How to know what test

* If you assume normality, it has to be **t** test.

* Terms like **under regular normality assumption**, or any hypothesis test about the **mean**.




## When nonparametric

* In this class, if the question says: **Test if the median of ... **

* Any terminology that says **without any parametric assumption**.

* If there is no keyword about parametric vs nonparametric, you are free to choose.


## Equivalence between tests


| Parametric test | Code| Non-parametric test | Code|
|---|---|---|---|
| One sample $t$ test | $\texttt{t.test}$ |?? | ??| 
| Two sample independent test| $\texttt{t.test}$| Wilcoxon rank-sum test| $\texttt{wilcox.test}$ |
| Two sample paired test| $\texttt{t.test}$| Wilcoxon signed-rank test| $\texttt{wilcox.test}$ |



## Unlearn something

* Let's simplify.
* Forget the following two options:
  -  $\texttt{var.equal=}$
  -  $\texttt{exact=}$



## One sample hypothesis test: parametric


My following code downloads the bodyfat data^[http://hbiostat.org/data] in $\texttt{R}$ in a dataframe named $\texttt{bodyfat_data}$. 

* Let's do the homework 1 question:
```{r}
 bodyfat_data = read.csv(url("https://hbiostat.org/data/repo/bodyfat.csv"))
```

Test the null hypothesis that population mean for the $\texttt{BodyFat}$ variable is less than $17$.



## Answer:


* Denote $\mu$ as the population mean for the $\texttt{BodyFat}$ variable.
* Hypotheses: $H_0$: $\mu \geq 17$; $H_1$: $\mu < 17$.
* One sample $t$ test.
* First, let's save the $\texttt{BodyFat}$ variable in a new vector and then write the rest of the code.

```{r}
bfat = bodyfat_data$BodyFat
```


## Cont.


```{r}
bfat = bodyfat_data$BodyFat
t.test(bfat, mu=17, alternative="less", conf.level=0.95)
```

Since $p$ value > $\alpha = 0.05$, can not reject $H_0$.



## One sample hypothesis test: nonparametric


* **Question 1**: Can you do a similar nonparametric test of the above problem? Specify in your hypothesis what test you are doing here and explain the symbols.

* **Question 2**: Under no parametric assumption, test that population median for the $\texttt{BodyFat}$ variable is less than $17$.

 
 
## One sample: nonparametric
 
 
 

* Denote $M$ as the population mean for the $\texttt{BodyFat}$ variable.
* Hypotheses: $H_0$: $M \geq 17$; $H_1$: $M < 17$.
* One sample test; so the **sign test**.


## Sign test

```{r}
bfat = bodyfat_data$BodyFat
positives_cnt = sum(bfat> 17)
negatives_cnt = sum(bfat< 17)
binom.test(positives_cnt, positives_cnt+negatives_cnt,
           alternative="less", conf.level = 0.95)
```



## Let's simplify

* If you are confused, **forget the above.**
* There is a version of Wilcoxon test that eases up the coding job.
* You copy the code for the t test.
* Then replace $\texttt{t.test}$ with $\texttt{wilcox.test}$.


## Wilcoxon test for one sample

```{r}
bfat = bodyfat_data$BodyFat
wilcox.test(bfat, mu=17, alternative="less", conf.level=0.95)
```





## Takeaway?


* The code is $\texttt{t.test}(\ldots)$ or $\texttt{wilcox.test}(\ldots)$.
* Put $\texttt{mu=}$ as the mean value <span style="color:red">IN THE HYPOTHESIS</span>.
* Use $\texttt{alternative="two.sided"}$; or $\texttt{alternative="greater"}$, or $\texttt{alternative="less"}$. 
* Use $\texttt{conf.level= 1 -} \alpha$.
* <span style="color:purple">Same syntax in parametric and nonparametric; but only change the name of the function</span>.





## Two sample hypothesis test


* Once again, if you are confused, 
  - FORGET $\texttt{var.equal=}$.
  - FORGET $\texttt{exact=}$.

* Let's do the parametric first; then replace $\texttt{t.test}$ with $\texttt{wilcox.test}$ for nonparametric.



## HW2 Question 1


* 16 students in a Statistics course are randomly divided into two groups of 8, each assigned to only one  of two test formats -- Project-based or Concept-based. The test scores for each group are as follows:
<center> **Project-based:** <span style = "color:green"> 38, 54, 63, 45, 40, 49, 58, 57 </center> 
<center> **Concept-based:** <span style = "color:green"> 55, 84, 79, 65, 78, 74, 58, 69 </center>



* Under regular normality assumptions, test whether the mean scores of the two tests are equal at $\alpha = 0.05$.

* Parametric/ nonparametric?
* One/ two-tailed?
* $\texttt{alternative=}$ ?


## Cont.

* Parametric since it says **under regular normality assumptions**.
* Two-tailed since it says test if the mean of two tests are equal. So,  we are testing **equal to** against **not equal to**.
* $\texttt{alternative= "two.sided"}$.

## In $\texttt{R}$.

* For independent test, **all values should be put in one vector first.**
 
```{r}
# First, all values in ONE vector
all_scores = c(38, 54, 63, 45, 40, 49, 58, 57,
               55, 84, 79, 65, 78, 74, 58, 69)
```

* Step 2: **All the groups in the second vector.**
```{r} 
 #Second, create a vector of test information
test_types = c(  rep("1.Project", 8),
                 rep("2.Concept", 8))
```


## Cont.

* For the test, you have to do:
```{r eval=F}
t.test(<value_vector> ~ <group_vector>,  mu = ..., 
    alternative = ..., conf.level = ...).
```

```{r}
t.test(all_scores ~ test_types, mu=0, 
       alternative="two.sided",   conf.level = 0.95)
```


## Cont.

* Look at the p value.
* In this case, reject $H_0$.

## For nonparametric?

* Wilcoxon rank sum test.
* Replace $\texttt{t.test}$ with $\texttt{sign.test}$.
* Question statement can say:
  
  - **Q1**: <span style="color:blue">Without assuming any parametric distribution</span>, test whether the median scores of the two test formats are equal at $\alpha = 0.05$. 
  - **Q2** Test whether the <span style="color:blue">median</span> scores of the two test formats are equal at $\alpha = 0.05$. 
  - **Q3**: Perform the <span style="color:blue">equivalent nonparametric test</span> for the above question. Explain notations and hypotheses.
  


## Cont: Nonparametric

```{r}
wilcox.test(all_scores ~ test_types,  mu=0, 
             alternative="two.sided", conf.level = 0.95)
```

* Enough evidence to reject $H_0$.






## HW 2 Q3


* Practice hypothesis testing with datasets. My following code downloads the Diabetes Data^[<https://hbiostat.org/data>, courtesy of the Vanderbilt University Department of Biostatistics] in $\texttt{R}$ in a dataframe named $\texttt{diabetes_data}$. You need to be connected to internet for this.

```{r}
diabetes_data = read.csv(url("https://hbiostat.org/data/repo/diabetes.csv"))
```


* Assuming normality, test whether there is a significant difference in the mean Stabilized Glucose levels (column name = **stab.glu**) between males and females (column name = **gender**). $\alpha = 0.10$.


## Cont.

* Independent since in this example, each subject has either gender = male or gender = female.
* Parametric since the question says **assuming normality**.
* Two-tailed since we test **whether there is a significant difference**, i.e., **difference = 0** vs **difference $\neq$ 0**.

## Cont.

* Step 1:  **All values should be put in one vector first.** Here all the values of glucose level are in **stab.glu** column.

```{r}
Glucose = diabetes_data$stab.glu   # Values
```


* Step 2: **All the groups in the second vector.**.

```{r}
Gender  = diabetes_data$gender     # Groups
```


## Cont.

* Step 3: Do the test.

```{r}
t.test(Glucose  ~ Gender, mu = 0, 
       alternative = "two.sided", conf.level = 1-0.10)
```

* Step 4: Finish the conclusion.

## Nonparametric equivalent questions

* Without any parametric assumption, test whether there is a significant difference in the median Stabilized Glucose levels (column name = **stab.glu**) between males and females (column name = **gender**). $\alpha = 0.10$.

* Do the nonparametric equivalent test for the last example.

* <span style="color:purple"> Just replace $\texttt{t.test}$ with $\texttt{wilcox.test}$.


## Cont.


```{r}
Glucose = diabetes_data$stab.glu   # Values
Gender  = diabetes_data$gender     # Groups
wilcox.test(Glucose  ~ Gender, mu = 0, 
       alternative = "two.sided", conf.level = 1-0.10)
```




## Takeaway?

* Write clearly first if it an <span style="color:purple">independent</span> or <span style="color:purple"> paired </span> test.

* If independent test:
  
  - First: **All values should be put in one vector first.**
  - Second: **All the groups in second vector.**

```{r eval=F}
t.test(<value_vector> ~ <group_vector>,  mu = ..., 
    alternative = ..., conf.level = ...).
```





## HW2 Q3A


* Do a nonparametric test that the median age (column name = **age**) is less than 40.5.


## Q3A

* What would you do if it were a t test/ parametric test?

* $H_0$: $\mu \geq 40.5$; $H_A$: $\mu < 40.5$.
```{r}
Ages = diabetes_data$age
t.test(Ages, mu=40.5,
       alternative="less", conf.level=0.90)
```


## HW3 Q3A

* Replace $\texttt{t.test}$ with $\texttt{wilcox.test}$.

```{r}
Ages = diabetes_data$age
wilcox.test(Ages, mu=40.5,
       alternative="less", conf.level=0.90)
```



## HW3 Q2A

All the 8 students from a Statistics class take both a project-based test and a concept-based test. The following are their scores for each test:

|           | Student 1 | Student 2 | Student 3 | Student 4 | Student 5 | Student 6 | Student 7 | Student 8 |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| **Project-based** |     38    |     54    |     63    |     45    |     40    |     49    |     58    |     57    |
| **Concept-based** |     55    |     84    |     79    |     65    |     78    |     74    |     58    |     69    |


## Cont.

A. Assuming normality, test if the mean population score for the Project-based test is smaller than that of the Concept-based test.


* Q. Paired/ Independent?
* Q. One or two sided?
* Q. $\texttt{alternative=}$?



## Cont.

* Paired since every subject has data in both samples.
* One-sided since question says "smaller".
* Q. $\texttt{alternative= "less"}$.

## Paired test

* For paired test:
  - Create a vector for the first sample of data. 
  - Create a second vector for the second sample of data. 

* Syntax:

```{r eval=F}
t.test(<vector1>, <vector2>,  paired=TRUE, mu = ...,
    alternative = ..., conf.level = ...).
```


## Cont.

```{r}
Project_based = c(38, 54, 63, 45, 40, 49, 58, 57)
Concept_based = c(55, 84, 79, 65, 78, 74, 58, 69)

t.test(Project_based, Concept_based, paired=TRUE,
       alternative = "less",
       mu=0, conf.level = 1-0.05)
```




## For nonparametric


* What if the question says one of the following:
  
  - Replicate the above test with an equivalent nonparametric one.
  - Test if the population median of the two tests are equal or not.
  - Under no parametric assumption, can you test $\ldots$?
  
* You do the nonparametric test: Wilcoxon signed-rank test.

* Replace $\texttt{t.test}$ with $\texttt{wilcox.test}$.


## Cont.

```{r}
Project_based = c(38, 54, 63, 45, 40, 49, 58, 57)
Concept_based = c(55, 84, 79, 65, 78, 74, 58, 69)

t.test(Project_based, Concept_based, paired=TRUE,
       alternative = "less",
       mu=0, conf.level = 1-0.05)
```





## Equivalence between tests


| Parametric test | Code| Non-parametric test | Code|
|---|---|---|---|
| One sample $t$ test | $\texttt{t.test}$ | Wilcoxon rank-sum test | $\texttt{wilcox.test}$| 
| Two sample independent test| $\texttt{t.test}$| Wilcoxon rank-sum test| $\texttt{wilcox.test}$ |
| Two sample paired test| $\texttt{t.test}$| Wilcoxon signed-rank test| $\texttt{wilcox.test}$ |

