---
title: "Nonparametric Tests"
subtitle: "One and Two Sample Tests"
author: "Ranadeep Daw"
date: "`r Sys.Date()`"
output:
  revealjs::revealjs_presentation:
    theme: solar
    center: true
    widescreen: true
    width: 1600
    height: 900
    df_print: kable
    reveal_options:
      slideNumber: true
      previewLinks: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Recap

1. Calculation of descriptive statistics, e.g., mean, median, quantiles, IQR.
2. Estimation under normality assumption.
3. Confidence intervals and hypothesis testing.
  - One sample $t$ test: Two-sided and one-sided
  - Two sample $t$ test: Two-sided and one-sided
    * Paired sample
    * Independent sample
      - Equal Variance
      - Unequal Variance
      
      
## Recap: $\texttt{R}$ Syntaxes

1. <span style = "color:blue;">Descriptive: </span> $\texttt{mean}(...)$, $\texttt{median}(...)$, $\texttt{quantile}(...)$, $\texttt{boxplot}(...)$.
2. <span style = "color:blue;">T-test and CI:</span> $$\texttt{t.test(..., alternate=..., mu=..., conf.level = ...)}$$
3. <span style = "color:blue;">Two sample T-test and CI:</span> $$\texttt{t.test(vec1, vec2, alternate=..., mu=..., conf.level = ...)}$$
  * <span style = "color:limegreen;">For paired sample, </span> add **$\texttt{paired=TRUE}$**.
  * <span style = "color:limegreen;">For independent sample, </span> **$\texttt{paired=FALSE}$** (by default).
    - <span style = "color:purple;">Equal variance:</span> add **$\texttt{vğšŠğš›.ğšğššğšğšŠğš• = FALSE}$**.
    - <span style = "color:purple;">Unequal variance:</span> **$\texttt{ğšŸğšŠğš›.ğšğššğšğšŠğš• = TRUE}$** (by default).



## Normality Assumptions

* Recall, we assumed that our sample is from a population that follows a normal distribution $\mathbb{N}( \mu, \sigma^2)$.

  - Symmetric around it's mean, population mean = population median = $\mu$.
  - High probabilities near the center, lower probabilities near the tails.
  
* What if our assumptions are wrong?

## Probability Distribution

```{r echo=F}
# Define mean and standard deviation
mu <- 0
sigma <- 1

# Generate a sequence of x values
x <- seq(mu - 4 * sigma, mu + 4 * sigma, length = 1000)

# Calculate the normal density
y <- dnorm(x, mean = mu, sd = sigma)

# Plot the normal distribution curve
plot(x, y, type = "l", lwd = 2, col = "blue", xaxt = "na",
     xlab = "x",
     ylab = "Density",
     main = "Normal Probabilities")

# Shade regions and add vertical lines
regions <- list(
  list(from = mu - 4*sigma, to = mu - 3 * sigma, col = "lightblue", label = "0.0015"),
  list(from = mu - 3 * sigma, to = mu - 2 * sigma, col = "lightblue", label = "0.02"),
  list(from = mu - 2 * sigma, to = mu - sigma, col = "lightgreen", label = "0.14"),
  list(from = mu - sigma, to = mu, col = "lightpink", label = "0.34"),
  list(from = mu + sigma, to = mu, col = "lightpink", label = "0.34"),
  list(from = mu + 2 * sigma, to = mu + sigma, col = "lightgreen", label = "0.14"),
  list(from = mu + 3 * sigma, to = mu + 2 * sigma, col = "lightblue", label = "0.02"),
  list(from = mu + 4 * sigma, to = mu + 3 * sigma, col = "lightblue", label = "0.0015")
)
  


# Add shaded regions and labels
for (region in regions) {
  rect(region$from, 0, region$to, dnorm(region$from), border = NA)
  text(mean(c(region$from, region$to)), dnorm(mu) * 0.1, labels = region$label, col = "black")
}

# Add vertical lines
abline(v = mu, lty = 2)
abline(v = c(mu - sigma, mu + sigma), lty = 2)
abline(v = c(mu - 2 * sigma, mu + 2 * sigma), lty = 2)
abline(v = c(mu - 3 * sigma, mu + 3 * sigma), lty = 2)

# Add axis labels using Greek letters
axis(1, at = c(mu - 4 * sigma, mu - 3 * sigma, mu - 2 * sigma, mu - sigma, mu, mu + sigma, mu + 2 * sigma, mu + 3 * sigma, mu + 4 * sigma),
     labels = expression(-infinity, mu - 3 * sigma, mu - 2 * sigma, mu - sigma, mu, mu + sigma, mu + 2 * sigma, mu + 3 * sigma, infinity)
     )


```



## Departure from Normality Assumptions

\newcommand{\X}{\mathrm{X}}



- **Departure from Normality**: Occurs when the sample suggests that the population does not follow a normal distribution.
- **Consequences**:
  - Can affect the validity of statistical tests (e.g., t-tests).
  - May lead to incorrect conclusions.
- **Common Causes of Departure:**
  - **Skewness**: Asymmetry in the distribution.
  - **Kurtosis**: Peakedness or flatness compared to a normal distribution.



## Example

```{r echo=F, out.width=700}
data = rnorm(5000, 0, 1)
par(mfrow = c(1, 2))
# Example R code to plot histogram
hist(data, main = "Example of Normality", xlab = "x", ylab = "Count", col = "skyblue", border = "black", breaks = 100)
hist(data[data > 0], main = "Departure from Normality", xlab = "x", ylab = "Count", col = "skyblue", border = "black", breaks = 100)
```



## QQ Plot

* **Definition:** Plots observed data quantiles against theoretical quantiles of a normal distribution.
* **Interpretation:** If the points approximately lie on a straight line, the data is normally distributed.

  - **Look for**: Departure from normality if points are not on a straight line.
* $\texttt{R}$ Syntax: 

```{r eval=F, out.width=500}
qqnorm(<vector>)
qqline(<vector>)
```

## Example: Evidence of normality

```{r echo=T, out.width=500}
# Example R code to plot Q-Q plot
data = rnorm(5000, 0, 1)
qqnorm(data)
qqline(data, col = "red")


```



## Example: Evidence of Departure from normality


```{r echo=T, out.width=500}
data2 = data[data < 0]
qqnorm(data2)
qqline(data2, col = "red")
```

## Non parametric Tests


- **Definition**: Statistical methods that do not assume a specific distribution for the data.
- **Contrast with Parametric Methods**:
  - **Parametric**: Assumes data follows a known distribution (e.g., normal distribution).
  - **Nonparametric**: Makes no such assumptions about the underlying data distribution. <span style="color:green">It still has parameters.</span>

## Examples of Usage of Nonparametric Methods

- **Small Sample Sizes**: When the sample size is too small to reliably estimate parameters.
- **Non-Normal Data**: When data significantly departs from normality.
- **Ordinal Data**: When dealing with ordinal data or ranks rather than interval data.

## One Sample Nonparametric Test: Sign Test

- **Sign Test**: A nonparametric test used to evaluate if the median of a single sample differs from a hypothesized value.
- **Purpose**: Test if the population median of a sample $\X = \{x_1, \ldots, x_n \}$ is $\mu_0$.
- ** Procedure**: Assign  a "plus" (+) to numbers $\geq$ $\mu_0$, a "minus" otherwise.
  i. If $x_i > \mu_0$, it's a +.
  ii. If $x_i < \mu_0$, it's a -.
  iii. Under $H_0$, the number of **+** and **-** should be roughly equal, following a binomial distribution with $p = 0.5$.

## Example: Median MPG in `mtcars` Dataset

- $H_0$: Median mpg = 22 vs $H_A$: Median mpg $\neq$ 22. Test at $\alpha = 0.05$.

```{r}
data(mtcars)
hypothised_median = 22
positives_count   = sum(mtcars$mpg > hypothised_median)
negatives_count   = sum(mtcars$mpg < hypothised_median)

sign_test <- binom.test(positives_count, positives_count + negatives_count, 
                               p = 0.5, alternative = "two.sided")
sign_test
```

* Decision: Since $p$ value < $\alpha$, enough evidence to reject $H_0$.




## Wilcoxon Sign-Rank Test

* Can be generalized to one or two sample tests.
* 