---
title: "Test for Variance"
subtitle: "Chi square and F tests"
author: "Ranadeep Daw"
date: "`r Sys.Date()`"
output:
  revealjs::revealjs_presentation:
    theme: "solar"
    center: true
    widescreen: true
    self_contained: true
    width: 1600
    height: 900
    df_print: kable
    reveal_options:
      slideNumber: true
      previewLinks: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```


## One sample Variance Test

* **Data**: $x_1, x_2, \ldots, x_n$.
* **Assumptions**: $X \sim \mathbb{N}(\mu, \sigma^2)$.
* $H_0: \sigma^2 = \sigma_0^2$ vs $H_A: \sigma^2 \neq, \leq, \geq  \sigma_0^2$.
* Sample variance: $s^2 = \frac{1}{n-1} \sum_{i=1}^{n} (x_i - \bar{x})^2$. 


## Procedure

* Test statistic: $T = \frac{(n-1) s^2}{ \sigma_0^2}$ 
* Reject $H_0$ if: 
  -  $T < \chi^2_{\alpha, n-1}$ if $H_A: \sigma^2 < \sigma_0^2$.
  -  $T > \chi^2_{1-\alpha, n-1}$ if $H_A: \sigma^2 > \sigma_0^2$.
  -  $T > \chi^2_{1-\alpha/2, n-1}$ and $T < \chi^2_{\alpha/2, n-1}$ if $H_A: \sigma^2 \neq \sigma_0^2$.
  

## Graphical Representation

```{r echo=F}
# Load necessary libraries
library(ggplot2)

# Chi-square distribution parameters
df <- 10  # degrees of freedom
alpha <- 0.05  # significance level

# Critical values for the three tests
left_critical <- qchisq(alpha, df)
right_critical <- qchisq(1 - alpha, df)
two_tail_lower <- qchisq(alpha / 2, df)
two_tail_upper <- qchisq(1 - alpha / 2, df)

# Create a sequence of x values (chi-square values)
x_vals <- seq(0, 30, length.out = 1000)

# Chi-square density values
y_vals <- dchisq(x_vals, df)

# Create the base plot
plot_base <- ggplot(data.frame(x = x_vals, y = y_vals), aes(x, y)) +
  geom_line(color = "blue") +
  theme_minimal() +
  labs(x = "Chi-Square Statistic", y = "Density", title = "Rejection Regions for Chi-Square Test")

# Left-tailed rejection region
left_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x < left_critical), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = left_critical, linetype = "dashed", color = "red") +
  annotate("text", x = left_critical - 2, y = 0.02, label = paste("Critical Value"), color = "red") +
  ggtitle("Left-Tailed Test: Rejection Region")

# Right-tailed rejection region
right_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x > right_critical), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = right_critical, linetype = "dashed", color = "red") +
  annotate("text", x = right_critical + 2, y = 0.02, label = paste("Critical Value"), color = "red") +
  ggtitle("Right-Tailed Test: Rejection Region")

# Two-tailed rejection region
two_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x < two_tail_lower), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x > two_tail_upper), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = two_tail_lower, linetype = "dashed", color = "red") +
  geom_vline(xintercept = two_tail_upper, linetype = "dashed", color = "red") +
  annotate("text", x = two_tail_lower - 1.5, y = 0.02, label = paste("Lower Critical Value"), color = "red") +
  annotate("text", x = two_tail_upper + 2, y = 0.02, label = paste("Upper Critical Value"), color = "red") +
  ggtitle("Two-Tailed Test: Rejection Region")

# Display the plots
library(gridExtra)
grid.arrange(left_tail_plot, right_tail_plot, two_tail_plot, ncol = 1)

```



## Example for $\texttt{mpg}$ from $\texttt{mtcars}$

* You will need to install the package $\texttt{EnvStats}$.

```{r eval=F}
install.packages("EnvStats")
```

* Let's do the test if variance of mpg is = 36 against less than 36.
* Syntax:

```{r eval=F}
varTest(x, alternative = "two.sided/less/greater", conf.level = ..., 
    sigma.squared = <value_of_the_hypotheses>)
```


## Example

```{r}
library("EnvStats")
data("mtcars")
MPG = mtcars$mpg
varTest(MPG, alternative = "less", conf.level = 0.95,  sigma.squared = 36, data.name = NULL)
```




## Two-sample equality of variance

* Data: 
  - $\mathbb{X}_1: \{x_{11}, x_{12}, \ldots, x_{1m} \}$
  - $\mathbb{X}_2: \{x_{21}, x_{22}, \ldots, x_{2n} \}$
  
* Assumption: $\mathbb{X}_1 \sim \mathbb{N}(\mu_1, \sigma^2_1)$, $\mathbb{X}_2 \sim \mathbb{N}(\mu_2, \sigma^2_2)$.
* Hypotheses: $H_0: \sigma_1^2 = \sigma_2^2$ vs $H_A: \sigma_1^2 \leq, \neq, \geq \sigma_2^2$.


## Procedure

* **Test Statistic:** $F = \frac{s_1^2}{s_2^2}$;  where \(s_1^2\) and \(s_2^2\) are the sample variances.
* Reject if:
  -  $F < F_{\alpha, m-1, n-1}$ if $H_A: \sigma_1^2 < \sigma_2^2$.
  -  $F > F_{1-\alpha, m-1, n-1}$ if $H_A: \sigma_1^2 > \sigma_2^2$.
  -  $F > F_{1-\alpha/2, m-1, n-1}$ or $F < F_{\alpha/2, m-1, n-1}$ if two sided test.


## $\texttt{R}$ Syntax

* Let's test if there is a difference in the variances of $\texttt{mpg}$ in the $\texttt{mtcars}$ data based on the transmission (Variable = $\texttt{am}$).

* Syntax:

```{r eval=F}
var.test(<var1> ~ <var2>, 
         alternative = "two.sided / less / greater", 
         conf.level = ...)
```

## Example

```{r}
data("mtcars")
MPG = mtcars$mpg
Automatic = mtcars$am
var.test(MPG ~ Automatic, alternative = "two.sided", conf.level = 0.95)
```


## Graphical Representation

```{r echo=F}
# Load necessary libraries
library(ggplot2)
library(gridExtra)

# Parameters for the F-distribution
df1 <- 10  # degrees of freedom for the numerator (group 1)
df2 <- 15  # degrees of freedom for the denominator (group 2)
alpha <- 0.05  # significance level

# Critical values for upper-tailed, lower-tailed, and two-tailed tests
upper_critical <- qf(1 - alpha, df1, df2)
lower_critical <- qf(alpha, df1, df2)
two_tail_lower <- qf(alpha / 2, df1, df2)
two_tail_upper <- qf(1 - alpha / 2, df1, df2)

# Create a sequence of F values for plotting
x_vals <- seq(0, 5, length.out = 1000)
y_vals <- df(x_vals, df1, df2)

# Base plot setup
plot_base <- ggplot(data.frame(x = x_vals, y = y_vals), aes(x, y)) +
  geom_line(color = "blue") +
  theme_minimal() +
  labs(x = "F Statistic", y = "Density", title = "Rejection Regions for F-Test")

# Upper-tailed rejection region
upper_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x > upper_critical), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = upper_critical, linetype = "dashed", color = "red") +
  annotate("text", x = upper_critical + 0.1, y = 0.02, label = paste("Critical Value"), color = "red") +
  ggtitle("Upper-Tailed Test: Rejection Region")

# Lower-tailed rejection region
lower_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x < lower_critical), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = lower_critical, linetype = "dashed", color = "red") +
  annotate("text", x = lower_critical - 0.2, y = 0.02, label = paste("Critical Value"), color = "red") +
  ggtitle("Lower-Tailed Test: Rejection Region")

# Two-tailed rejection region
two_tail_plot <- plot_base +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x < two_tail_lower), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_area(data = subset(data.frame(x = x_vals, y = y_vals), x > two_tail_upper), aes(x = x, y = y), fill = "red", alpha = 0.5) +
  geom_vline(xintercept = two_tail_lower, linetype = "dashed", color = "red") +
  geom_vline(xintercept = two_tail_upper, linetype = "dashed", color = "red") +
  annotate("text", x = two_tail_lower - 0.2, y = 0.02, label = paste("Lower Critical"), color = "red") +
  annotate("text", x = two_tail_upper + 0.1, y = 0.02, label = paste("Upper Critical"), color = "red") +
  ggtitle("Two-Tailed Test: Rejection Region")

# Display the plots
grid.arrange(upper_tail_plot, lower_tail_plot, two_tail_plot, ncol = 1)

```

