---
title: "Multiple Linear Regression"
subtitle: "Examples using R"
author: "Ranadeep Daw"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 2
    keep_md: false
---
  
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```



# Question 1: MLR


### Let's again go back to the OzDASL website to download the **Cheese Taste** data. Metadata link is [here](<https://lib.stat.cmu.edu/DASL/Datafiles/Cheese.html>). Following code downloads the same data from a different site in a dataframe called $\texttt{cheese_data}$ that has the data from the above link.


```{r}
cheese_data = read.csv(url("https://raw.githubusercontent.com/dawranadeep/24FA_git/refs/heads/main/Practice_qns/img/cheese_data.csv"))
```

### Look how the first few rows of the data look like:

```{r}
head(cheese_data)
```


### <br> A. Draw all the scatterplots between the response variable **taste** and the other predictors. Any comments on the plots?

**Answer:** 

```{r}
# Code
```


### <br> B. Build the multiple linear regression model with **taste** as the response and other variables as predictors. Write down the fitted equation.



**Answer:** 

```{r}
# Code
```


### <br> C. What are the $R^2$ and adjusted $R^2$? Is this model significant?



**Answer:** 

```{r}
# Code
```




### <br> D. Test for $H_0: \beta_1 = 0$ vs $H_A: \beta_1 \neq 0$. Take $\alpha = 0.05$.

**Answer:** 

```{r}
# Code
```


### <br> E. Test for $H_0: \beta_2 = 0$ vs $H_A: \beta_2 \neq 0$. Take $\alpha = 0.05$.

**Answer:** 

```{r}
# Code
```


### <br> F. Find the $95\%$ confidence intervals of the estimated coefficients. Can you perform the tests in part D and E from this CI?

**Answer:** 

```{r}
# Code
```


### <br> G. Do the residual-vs-fitted plot. Can you comment anything on the residual?


**Answer:** 

```{r}
# Code
```


### <br> H. Do the QQ-plot of the residual. Can you comment anything about normality?


**Answer:** 

```{r}
# Code
```



### <br> I. Let's try to do backward selction using the $p$ values. How would you update the model? 

**Answer:** 

```{r}
# Code
```




### <br> J. Now do the backward selction using the $\texttt{step}$ function from $\texttt{R}$. What is the final model? 

**Answer:** 

```{r}
# Code
```


### <br> K. Comparing adjusted $R^2$, which model would you prefer?

**Answer:** 



<br><br><br>

# Question 2: Polynomial Regression and Model Selection

### Let's again go back to the OzDASL website to download the **Sound Pressure of Dolphin Sonar** data. Metadata link is [see here](<http://www.statsci.org/data/general/dolphin.html>). Following code downloads the data in a dataframe called $\texttt{dolphin_data}$ that has the data from the above link.

```{r}
dolphin_data = read.table(url("http://www.statsci.org/data/general/dolphin.txt"), header = TRUE)
```





### <br> A. Draw the scatterplot. What kind of model do you think you can build here?

**Answer:** 

```{r}
# Code
```



### <br> B. First, build a linear regression model. What are the $R^2$ and adjusted $R^2$? Is this model significant?


**Answer:** 

```{r}
# Code
```


### <br> C. Build a quadratic regression model now. What are the $R^2$ and adjusted $R^2$? Is this model significant?



**Answer:** 

```{r}
# Code
```


### <br> D. Comparing by adjusted $R^2$ values, which model would you prefer?



**Answer:** 



### <br> E. Let's try to do backward selction using the $p$ values. How would you update the model? Looking at the parameter significance values, which model would you prefer?



**Answer:** 




### <br> F. Now, try the $\texttt{step}$ function to do backward selction using AIC. Which model is your $\texttt{R}$ selecting?



**Answer:** 



### <br> G. Let's take the quadratic model you have done so far. Can you check if the residuals follow a normal distribution?


**Answer:** 


```{r}
# Code
```


### <br> H. For the above, please plot the fitted vs residual plot. What comments can you make from this plot?


**Answer:** 


```{r}
# Code
```





### <br> I. Let's take the linear regression model. Can you check if the residuals follow a normal distribution?


**Answer:** 


```{r}
# Code
```


### <br> H. Again, take the linear regression model and draw the residual-vs-fitted plot. What comments can you make from this plot?


**Answer:** 

```{r}
# Code
```

