---
title: "Writing with R Markdown"
subtitle: "Practice Set"
author: "Your Name Here"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
always_allow_html: true
---


```{r include = FALSE}
knitr::opts_chunk$set(error = T,  echo =  T, eval = T, warning =T, message = T)
```


# Notes: {-}

0. Download $\texttt{R}$ (https://cran.rstudio.com) and $\texttt{RStudio}$ (https://posit.co/download/rstudio-desktop/) in your computer. Alternatively, access $\texttt{RStudio}$ at <https://rstudio.hmcse.uwf.edu>. 

1. Download this file from $\textit{Canvas}$ and open it with $\texttt{RStudio}$.

2. This will be the typical homework format for all the homework assignments in this semester. We are using  $\texttt{R}$ Markdown here, which allows us to type answer as well as display our computation using $\texttt{R}$. Use this assignment to practice writing Markdown.

3. First, we will learn how to author a document:
  a. See the block at the top of the document between the $``---"$ and $``---"$ syntax. This block is called **YAML** block.
  b. Write your name beside the ''author'' option above.
  c. If you want, you can give any title/ subtitle beside the "title" and "subtitle" options.
  d. I fixed the date as the current system date. No need to change this.

4. Write your answer after my questions. I have marked it with "Write your answer here". 
5. If your answer needs any $\texttt{R}$ computation, I have created the $\texttt{R}$ code chunks. Alternatively, you can also add a code chunk with the the keyboard shortcut **Ctrl + Alt + I** (Windows) or **Cmd + Option + I** (Mac) or the **Add Chunk** command in the editor toolbar.

6. After writing one answer, please click the **Knit** button, or **Ctrl+Shift+K** (Windows), or **Cmd+Shift+K** (Mac), to compile the document.

7. **Important: Don't wait too long to knit. If there is a mistake, you will find it as soon as it happens.**

8. Now, use the following question to practice writing Markdown.

<br>
<br>



# Authoring 
### Update the document author deatils. Hint: please go to the top of the document (YAML block), find the "author" field, and insert your name within "..." beside it. {-}

Your answer should be at the top of this document.

# Writing plain text.

### a. What is your major? {-}

Write your answer here.

### b. What is your learning goal from this course? {-}

Write your answer here.


# Writing $\texttt{R}$ code.

### a. Using $\texttt{R}$, sum any two numbers that you want and show the output. I have given you the sample code below. Make sure to remove ''#'' from the code. {-}

Write your answer within the $\texttt{R}$ code chunk. 
```{r}
# x = <<Any number you like>>
# y = <<Any number you like>>
# x + y
```




### b. Load the $\texttt{tidyverse}$ package in $\texttt{R}$. {-}

Write your answer within the $\texttt{R}$ code chunk.
```{r}
# Write here. Remember to remove "#"
# Remember, to load a package, you should use  library("...").
# If it shows an error in loading (there is no package called ...), the package is not installed. Then you have to install it first using  install.package("...").

```




## Practice everything together

- Now answer the next questions for the $\texttt{mtcars}$ data.

## Q1 - Q2: 
How many observations and variables do we have?

**Ans:** 
```{r}
# Your code here.
```

There are ... cars and ... variables in the data.


## Q3:
What are the variables (or column names)?
```{r}
# Your code here
```

The column names are ...

## Q6: Plot the "mpg" column against "disp" column?
```{r}
# Your code here
```


## Final Instructions

1. Don't forget to knit. 
2. Download the HTML file. This is what you will submit for homework.
